---
title: "Racial and ethnic disparities in data exclusion are present at baseline, but not at the 2-year follow up:"
output: html_document
date: "2025-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Next, we assessed the role of race/ethnicity in predicting exclusion due to high-motion status at baseline using both frequentist and Bayesian approaches. At baseline, exclusion due to high-motion rates on the full sample (n = 8561) were highest for Black participants (48.3%), followed by Hispanic (44.8%), Asian (43.5%), Other (37.9%) and White (35.4%) participants (Fig. 3, top). At the 2-year follow-up (n = 5789), overall exclusion rates declined across all groups, consistent with decreased motion across child development as well as increased exclusion and loss to followup for high-motion individuals. Rates remained highest among Black (33.5%) and Asian (29.7%) participants, followed by Other (28.1%), Hispanic (26.3%), and White (24%) participants (Fig. 3, bottom). Demographic breakdowns for the full sample at baseline and follow up are described in Table 5 and Table 6, respectively.** 

## 1. load required packages:

```{r}
required_packages <- c(
  "lme4", "dplyr", "broom", "bayestestR", "ggplot2", "stats", "e1071", "tidyr",
  "patchwork", "multcomp", "lmerTest", "broom.mixed", "tableone", "flextable",
  "officer", "readr", "jsonlite", "BayesFactor", "MuMIn", "brms","consort")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    cat("Installing", pkg, "...\n")
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

cat("✓ All packages loaded successfully\n")
```

## 2. configuration for data and output directories: 

```{r}
# Configuration constants
CONFIG <- list(
  motion_threshold = 0.2,
  na_codes = c("777", "888", "999"),
  na_numeric = c(777, 888, 999), # 777 refuse to answer 888 decline to answer/question not asked 999 dont know
  data_dir = "/Users/ricard/Desktop/poldrack/abcd/data/abcd-data-release-5/core/",
  dairc_data <- readRDS("/Users/ricard/Desktop/poldrack/projects/head-motion-exclusion/data/df_final_all-refactored.rds"),
  output_dir = "/Users/ricard/Desktop/poldrack/projects/head-motion-exclusion/outputs",
  data_path = "/Users/ricard/Desktop/poldrack/projects/head-motion-exclusion/data/df_final_included-refactored.rds",
  model_control = glmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 2e5)
  )
)

# Create output directory if needed
if (!dir.exists(CONFIG$output_dir)) {
  dir.create(CONFIG$output_dir, recursive = TRUE)
}

# Helper function to read ABCD CSV files
read_abcd_csv <- function(path, ...) {
  if (!file.exists(path)) {
    stop("File not found: ", path)
  }
  read_csv(path, na = CONFIG$na_codes, show_col_types = FALSE, ...)
}
```

## 3. helper functions

```{r}
# Helper function to prepare timepoint data (eliminates code duplication)
prepare_timepoint_data <- function(df, timepoint_name, timepoint_filter) {
  cat("Preparing", timepoint_name, "data...\n")
  
  initial_data <- df %>% filter(eventname == timepoint_filter)
  initial_n <- nrow(initial_data)
  
  # Check if 'exclude' column exists, if not use 'exclude_motion'
  exclude_col <- if("exclude" %in% names(initial_data)) "exclude" else "exclude_motion"
  
  processed_data <- initial_data %>%
    dplyr::select(src_subject_id, income, parental_education, sex, race_ethnicity_label, 
                  all_of(exclude_col), rel_family_id, rsfmri_meanmotion, eventname, 
                  site, age, imgincl_rsfmri_include) %>%
    rename(exclude = all_of(exclude_col)) %>%
    mutate(
      # Convert scale() matrices to numeric vectors (fixes model issues)
      income_z = as.numeric(scale(income)),
      parent_ed_z = as.numeric(scale(parental_education)),
      age_z = as.numeric(scale(age)),
      
      # Create clear binary coding
      female = ifelse(sex == 2, 1, 0),
      
      # Ensure proper factor coding
      race_ethnicity_label = as.factor(race_ethnicity_label),
      site = as.factor(site)
    )
  
  # Remove missing data with reporting
  complete_data <- processed_data %>% drop_na()
  final_n <- nrow(complete_data)
  removed_n <- initial_n - final_n
  
  cat("✓", timepoint_name, ":", initial_n, "→", final_n, 
      "participants (", removed_n, "removed due to missing data)\n")
  
  return(complete_data)
}

# Helper function to check model convergence
check_model_convergence <- function(model, model_name) {
  convergence_code <- model@optinfo$conv$opt
  if (convergence_code != 0) {
    warning(paste(model_name, "did not converge (code:", convergence_code, ")"))
    return(FALSE)
  } else {
    cat("✓", model_name, "converged successfully\n")
    return(TRUE)
  }
}

# Helper function to perform model comparison
perform_model_comparison <- function(full_model, reduced_model, comparison_name) {
  cat("\n=== ", comparison_name, " ===\n")
  
  # Likelihood ratio test
  lrt_results <- anova(full_model, reduced_model)
  print(lrt_results)
  
  # Extract results
  chi_sq <- lrt_results$Chisq[2]
  df_diff <- lrt_results$Df[2]
  p_value <- lrt_results$`Pr(>Chisq)`[2]
  
  cat("\nLRT Results:\n")
  cat("χ²(", df_diff, ") = ", round(chi_sq, 2), ", p = ", round(p_value, 4), "\n", sep = "")
  
  # Bayesian comparison
  tryCatch({
    bf_results <- bayesfactor_models(full_model, reduced_model)
    bf_value <- as.numeric(bf_results$BF[1])
    cat("Bayes Factor: BF₁₀ = ", round(bf_value, 2), "\n", sep = "")
    
    return(list(p_value = p_value, bf_value = bf_value, chi_sq = chi_sq, df = df_diff))
  }, error = function(e) {
    cat("⚠️ Bayesian comparison failed:", e$message, "\n")
    return(list(p_value = p_value, chi_sq = chi_sq, df = df_diff))
  })
}

# Helper function to create odds ratio data for plotting
extract_odds_ratios <- function(model, model_name) {
  tryCatch({
    or_data <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
      filter(grepl("race_ethnicity_label", term)) %>%
      mutate(
        race_ethnicity_label = gsub("race_ethnicity_label", "", term),
        model = model_name
      )
    
    # Select only the columns we need
    or_data <- or_data %>%
      dplyr::select(race_ethnicity_label, estimate, conf.low, conf.high, model)
    
    return(or_data)
  }, error = function(e) {
    cat("⚠️ Failed to extract odds ratios for", model_name, ":", e$message, "\n")
    return(data.frame())
  })
}
```

## 4. data loading and preparation

```{r data-loading}
cat("=== Loading ABCD Data ===\n")

# Validate file exists
if (!file.exists(CONFIG$data_path)) {
  stop("Data file not found: ", CONFIG$data_path)
}

# Load main dataset
df_final_included <- readRDS(CONFIG$data_path)
cat("✓ Loaded", nrow(df_final_included), "total records\n")

# Prepare both timepoints using consistent function
baseline_data <- prepare_timepoint_data(
  df_final_included, 
  "baseline", 
  "baseline_year_1_arm_1"
)

followup_data <- prepare_timepoint_data(
  df_final_included,
  "followup", 
  "2_year_follow_up_y_arm_1"
)

# Create longitudinal dataset
common_subject_ids <- intersect(baseline_data$src_subject_id, followup_data$src_subject_id)
cat("✓", length(common_subject_ids), "subjects have both timepoints\n")

longitudinal_data <- bind_rows(
  baseline_data %>% 
    filter(src_subject_id %in% common_subject_ids) %>% 
    mutate(timepoint = "baseline"),
  followup_data %>% 
    filter(src_subject_id %in% common_subject_ids) %>% 
    mutate(timepoint = "followup")
)

# Add follow-up indicator to baseline data
baseline_data <- baseline_data %>%
  mutate(has_followup_data = src_subject_id %in% followup_data$src_subject_id)

cat("✓ Created longitudinal dataset:", nrow(longitudinal_data), "observations\n")
cat("✓", sum(baseline_data$has_followup_data), "baseline subjects have followup\n")
```

## 5. Baseline exclusion analysis: 

**To identify predictors of data exclusion due to high head motion in the baseline data, we fitted a series of generalized linear mixed-effects models modeling exclusion based on high-motion status as a function of racial/ethnic group with covariates including household income, parental education, age, and sex while also accounting for random intercepts for study site and family ID. At baseline (ages 9-10), there were significant racial/ethnic disparities in exclusion based on high-motion status. Consistent with prior work examining high-head motion status (Cosgrove et al., 2022), a model comparison showed that including race/ethnicity significantly outperformed the model without race/ethnicity (χ²(4) = 21.13, p < .001); here again the Bayes Factor preferred the reduced model (BF10 = 5.2e-04), suggesting that the significant effect nonetheless provides weak evidence for the alternative hypothesis. Group-specific estimates revealed that, relative to White participants, the odds of being excluded for high head motion were significantly higher for Black (OR = 1.23, 95% CI [1.04, 1.45], p = .013), Hispanic (OR = 1.34, 95% CI [1.16, 1.56], p < .001), and Asian (OR = 1.50, 95% CI [1.07, 2.12], p = .02) participants. This was not the case for participants who identified as ‘Other’ (OR = 1.03, CI [0.88, 1.21], p = 0.71). These effects are illustrated in Fig. 4 in terms of predicted probabilities (probability of exclusion), adjusted for all continuous covariates at their mean values for race/ethnicity and mean head motion.**

```{r}

# Full model with race/ethnicity
baseline_full_model <- glmer(
  exclude ~ income_z + parent_ed_z + race_ethnicity_label + female + age_z +
    (1 | rel_family_id) + 
    (1 | site),
  data = baseline_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Reduced model without race/ethnicity
baseline_reduced_model <- glmer(
  exclude ~ income_z + parent_ed_z + female + age_z +
    (1 | rel_family_id) +
    (1 | site),
  data = baseline_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Check convergence
baseline_full_converged <- check_model_convergence(baseline_full_model, "Baseline Full Model")

baseline_reduced_converged <- check_model_convergence(baseline_reduced_model, "Baseline Reduced Model")


```

# model comparison

```{r}
# Model comparison
if (baseline_full_converged && baseline_reduced_converged) {
  baseline_results <- perform_model_comparison(
    baseline_full_model, 
    baseline_reduced_model,
    "Baseline Model Comparison"
  )
}

cat("\nBaseline Analysis Summary:\n")

cat("Sample size:", nrow(baseline_data), "participants\n")

bf_result_race = bayesfactor_models(baseline_reduced_model, baseline_full_model)
print(bf_result_race)

```

## summary of full baseline model: 
```{r}
summary(baseline_full_model)
```

## summary of reduced baseline model: 
```{r}
summary(baseline_reduced_model)
```

```{r}
baseline_fixed <- broom.mixed::tidy(baseline_full_model, 
                                   effects = "fixed", 
                                   conf.int = TRUE)

for(i in 1:nrow(baseline_fixed)) {
  cat(sprintf("%s: β = %.3f, 95%% CI [%.3f, %.3f], p = %.3f\n",
              baseline_fixed$term[i],
              baseline_fixed$estimate[i],
              baseline_fixed$conf.low[i], 
              baseline_fixed$conf.high[i],
              baseline_fixed$p.value[i]))
}
```

## 6. 2 year follow up exclusion models:

**In the 2-year follow-up data (ages 11-13), these racial/ethnic disparities in high-head motion status exclusion were no longer present. The GLMER analysis at the 2-year follow up revealed no significant differences in the odds of exclusion based on high-head motion status for any racial/ethnic group compared to White participants (χ²(4) = 2.5, p = 0.64, BF10 = 1.04e-07, in favor of the reduced model). **

```{r}
cat("=== Follow-up Motion Exclusion Analysis ===\n")

# Full model with race/ethnicity
followup_full_model <- glmer(
  exclude ~ income_z + parent_ed_z + race_ethnicity_label + female + age_z +
    (1 | rel_family_id) +
    (1 | site),
  data = followup_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Reduced model without race/ethnicity
followup_reduced_model <- glmer(
  exclude ~ income_z + parent_ed_z + female + age_z +
    (1 | rel_family_id) +
    (1 | site),
  data = followup_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Check convergence
followup_full_converged <- check_model_convergence(followup_full_model, "Follow-up Full Model")
followup_reduced_converged <- check_model_convergence(followup_reduced_model, "Follow-up Reduced Model")
```

# summary of full followup model:
```{r}
summary(followup_full_model)
```

# summary of reduced model followup model: 
```{r}
summary(followup_reduced_model)
```

# model comparison: 
```{r}
# Model comparison
if (followup_full_converged && followup_reduced_converged) {
  followup_results <- perform_model_comparison(
    followup_full_model, 
    followup_reduced_model,
    "Follow-up Model Comparison"
  )
}

cat("\nFollow-up Analysis Summary:\n")
cat("Sample size:", nrow(followup_data), "participants\n")

```

## Bayes factor for 2 yr follow up model:  
```{r}
bf_result_race = bayesfactor_models(followup_full_model, followup_reduced_model,
                                    denominator = followup_reduced_model)
print(bf_result_race)
```

## 7. Longtiudinal exclusion model analysis (n = 4658)

**A longitudinal model examining youth with data at both timepoints (n = 4658) simultaneously found that after controlling for covariates (sex, age, parental education, household income), race/ethnicity was not a statistically significant predictor of high-motion status. A model comparison confirmed that adding race/ethnicity did not significantly improve model fit (χ²(4) = 5.81, p = 0.214, BF10 = 2.10e-07 in favor of the reduced model), suggesting no overall effect of race/ethnicity across the two timepoints (Table 4). However, there was a significant effect of study visit on high-motion status, with participants more likely to be excluded at the baseline visit compared to the two-year follow-up (β = 0.99, z = 17.35, p < .001, BF10 = 1.39e+72 favoring the full model), consistent with a reduction in head motion across development and greater failure to retain individuals with higher trait levels of head motion.** 

```{r}

cat("=== Longitudinal Motion Exclusion Analysis ===\n")
# Full model with race/ethnicity
longitudinal_full_model <- glmer(
  exclude ~ income_z + parent_ed_z + female + race_ethnicity_label + eventname + age_z +
    (1 | rel_family_id/src_subject_id) +
    (1 | site),
  data = longitudinal_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Reduced model without race/ethnicity
longitudinal_reduced_model <- glmer(
  exclude ~ income_z + parent_ed_z + female + eventname + age_z +
    (1 | rel_family_id/src_subject_id) +
    (1 | site),
  data = longitudinal_data, 
  family = binomial,
  control = CONFIG$model_control
)

# Check convergence
long_full_converged <- check_model_convergence(longitudinal_full_model, "Longitudinal Full Model")
long_reduced_converged <- check_model_convergence(longitudinal_reduced_model, "Longitudinal Reduced Model")

# Model comparison
if (long_full_converged && long_reduced_converged) {
  longitudinal_results <- perform_model_comparison(
    longitudinal_full_model, 
    longitudinal_reduced_model,
    "Longitudinal Model Comparison"
  )
}

cat("Sample size:", length(common_subject_ids), "subjects (", nrow(longitudinal_data), "observations)\n")
```

# summary of longitudinal full model: 
```{r}
summary(longitudinal_full_model)
```

# summary of longitudinal reduced model: 
```{r}
summary(longitudinal_reduced_model)
```

## bayes factor for longtiduinal model: 
```{r}
bf_result_longitudinal = bayesfactor_models(longitudinal_full_model, longitudinal_reduced_model,
                                            denominator = longitudinal_reduced_model)
print(bf_result_longitudinal)
```


## 8. plotting exclusion at baseline:  

**At baseline, exclusion due to high-motion rates on the full sample (n = 8561) were highest for Black participants (48.3%), followed by Hispanic (44.8%), Asian (43.5%), Other (37.9%) and White (35.4%) participants (Fig. 3, top). At the 2-year follow-up (n = 5789), overall exclusion rates declined across all groups, consistent with decreased motion across child development as well as increased exclusion and loss to followup for high-motion individuals.**

```{r fig.height=8, fig.width=10}
# Calculate exclusion summaries
exclusion_summary_baseline <- baseline_data %>%  
  group_by(race_ethnicity_label) %>%
  summarise(
    n = n(),
    n_excluded = sum(exclude == 1, na.rm = TRUE),
    percent_excluded = 100 * n_excluded / n,
    .groups = "drop"
  )

exclusion_summary_followup <- followup_data %>%  
  group_by(race_ethnicity_label) %>%
  summarise(
    n = n(),
    n_excluded = sum(exclude == 1, na.rm = TRUE),
    percent_excluded = 100 * n_excluded / n,
    .groups = "drop"
  )

# Set consistent factor levels
race_levels <- c("White", "Black", "Hispanic", "Asian", "Other")
exclusion_summary_baseline$race_ethnicity_label <- factor(exclusion_summary_baseline$race_ethnicity_label, levels = race_levels)
exclusion_summary_followup$race_ethnicity_label <- factor(exclusion_summary_followup$race_ethnicity_label, levels = race_levels)

# Create baseline exclusion plot
plot_baseline_exclusion <- ggplot(exclusion_summary_baseline, aes(x = race_ethnicity_label, y = percent_excluded, fill = race_ethnicity_label)) +
  geom_col(show.legend = FALSE, width = 0.8, color = "black") +
  geom_text(aes(label = sprintf("%.1f%%", percent_excluded)), vjust = -0.5, size = 5, color = "black") +
  labs(
    title = "Baseline",
    x = "Race/Ethnicity",
    y = "Percentage Excluded (%)"
  ) +
  ylim(0, 75) + 
  scale_fill_viridis_d(option = "D") +
  theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.line = element_line(color = "black")
  )

# Create follow-up exclusion plot
plot_followup_exclusion <- ggplot(exclusion_summary_followup, aes(x = race_ethnicity_label, y = percent_excluded, fill = race_ethnicity_label)) +
  geom_col(show.legend = FALSE, width = 0.8, color = "black") +
  geom_text(aes(label = sprintf("%.1f%%", percent_excluded)), vjust = -0.5, size = 5, color = "black") +
  labs(
    title = "2-Year Follow-up",
    x = "Race/Ethnicity",
    y = "Percentage Excluded (%)"
  ) +
  ylim(0, 75) +
  scale_fill_viridis_d(option = "D") +
  theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.line = element_line(color = "black")
  )

# Extract odds ratios for plotting
if (baseline_full_converged) {
  or_baseline <- extract_odds_ratios(baseline_full_model, "Baseline")
  or_baseline$race_ethnicity_label <- factor(or_baseline$race_ethnicity_label, levels = race_levels[-1])
}

if (followup_full_converged) {  
  or_followup <- extract_odds_ratios(followup_full_model, "Follow-up")
  or_followup$race_ethnicity_label <- factor(or_followup$race_ethnicity_label, levels = race_levels[-1])
}

# Create odds ratio plots
if (exists("or_baseline") && nrow(or_baseline) > 0) {
  plot_or_baseline <- ggplot(or_baseline, aes(x = estimate, y = race_ethnicity_label)) +
    geom_point(color = "purple", size = 3) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "black") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "brown") +
    geom_text(aes(label = sprintf("%.2f", estimate)), vjust = -1, size = 5, color = "black") +
    scale_x_continuous(limits = c(0.5, 2.5), breaks = seq(0.5, 2.5, 0.5)) +
    labs(
      title = "Baseline",
      x = "Odds Ratio",
      y = "Race/Ethnicity"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.line = element_line(color = "black")
    )
}

if (exists("or_followup") && nrow(or_followup) > 0) {
  plot_or_followup <- ggplot(or_followup, aes(x = estimate, y = race_ethnicity_label)) +
    geom_point(color = "purple", size = 3) +  
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "black") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "brown") +
    geom_text(aes(label = sprintf("%.2f", estimate)), vjust = -1, size = 5, color = "black") +
    scale_x_continuous(limits = c(0.5, 2.5), breaks = seq(0.5, 2.5, 0.5)) +
    labs(
      title = "2-Year Follow-up", 
      x = "Odds Ratio",
      y = "Race/Ethnicity"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.line = element_line(color = "black")
    )
}

# Combine plots
if (exists("plot_or_baseline") && exists("plot_or_followup")) {
  combined_exclusion_plot <- (plot_baseline_exclusion | plot_followup_exclusion) / 
                            (plot_or_baseline | plot_or_followup) +
    plot_annotation(
      title = "fMRI Exclusion by Race/Ethnicity: Prevalence and Odds Ratios",
      theme = theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5))
    )
  
  print(combined_exclusion_plot)
  
  # Save plot
  ggsave(
    filename = "exclusion_oddsratio.svg",
    plot = combined_exclusion_plot,
    path = CONFIG$output_dir,
    width = 12,
    height = 10,
    units = "in"
  )
}

```

## create demographic tables:

```{r}
cat("=== Creating Demographic Tables ===\n")

# Prepare variables for tables
demographic_vars <- c("age_z", "sex", "race_ethnicity_label", "income", "parental_education")
categorical_vars <- c("sex", "race_ethnicity_label", "income", "parental_education")

# Convert age back to numeric for table
baseline_data$age_numeric <- as.numeric(baseline_data$age_z)
followup_data$age_numeric <- as.numeric(followup_data$age_z)

# Create baseline demographics table
baseline_table <- CreateTableOne(
  vars = c("age_numeric", demographic_vars[-1]), 
  strata = "exclude", 
  data = baseline_data, 
  factorVars = categorical_vars
)

print(baseline_table, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, test = TRUE)

```

# print 2yr follow up table: 
```{r}
# Create follow-up demographics table  
followup_table <- CreateTableOne(
  vars = c("age_numeric", demographic_vars[-1]), 
  strata = "exclude", 
  data = followup_data, 
  factorVars = categorical_vars
)

print(followup_table, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, test = TRUE)
```

# export tables to word doc: 
```{r}
# Export tables to Word documents
tryCatch({
  # Baseline table
  baseline_table_matrix <- print(baseline_table, showAllLevels = TRUE, quote = FALSE, 
                                noSpaces = TRUE, printToggle = FALSE, test = TRUE)
  baseline_table_df <- as.data.frame(baseline_table_matrix)
  baseline_flextable <- flextable(baseline_table_df)
  
  save_as_docx(
    baseline_flextable,
    path = file.path(CONFIG$output_dir, "baseline_demographics_table.docx")
  )
  
  # Follow-up table
  followup_table_matrix <- print(followup_table, showAllLevels = TRUE, quote = FALSE, 
                                noSpaces = TRUE, printToggle = FALSE, test = TRUE)
  followup_table_df <- as.data.frame(followup_table_matrix)
  followup_flextable <- flextable(followup_table_df)
  
  save_as_docx(
    followup_flextable,
    path = file.path(CONFIG$output_dir, "followup_demographics_table.docx")
  )
  
  cat("✓ Demographic tables saved to", CONFIG$output_dir, "\n")
  
}, error = function(e) {
  cat("⚠️ Failed to save demographic tables:", e$message, "\n")
})
```


## 9. consort flow diagram
```{r consort-flow-diagrams, fig.height=12, fig.width=10}

#dairc_data loaded in only for the purposes of this consort diagram. 
dairc_data_baseline <- prepare_timepoint_data(
  dairc_data, 
  "baseline", 
  "baseline_year_1_arm_1"
)

dairc_data_followup <- prepare_timepoint_data(
  dairc_data,
  "followup", 
  "2_year_follow_up_y_arm_1"
)


cat("=== Creating CONSORT Flow Diagrams ===\n")

baseline_data_n = nrow(baseline_data)
followup_data_n = nrow(followup_data)

dairc_data_full <- sum(dairc_data$eventname == "baseline_year_1_arm_1" | 
                       dairc_data$eventname == "2_year_follow_up_y_arm_1", na.rm = TRUE)
# Calculate flow numbers from dairc_data (from another script)
baseline_raw_n <- sum(dairc_data$eventname == "baseline_year_1_arm_1", na.rm = TRUE)
followup_raw_n <- sum(dairc_data$eventname == "2_year_follow_up_y_arm_1", na.rm = TRUE)

# Calculate DAIRC = 1 counts from dairc_data (these should be 10,280 and 6,538)
baseline_dairc_n <- sum(dairc_data$eventname == "baseline_year_1_arm_1" & 
                       dairc_data$imgincl_rsfmri_include == 1, na.rm = TRUE)

followup_dairc_n <- sum(dairc_data$eventname == "2_year_follow_up_y_arm_1" & 
                       dairc_data$imgincl_rsfmri_include == 1, na.rm = TRUE)

# Calculate exclusions for baseline analysis
baseline_excluded_motion <- sum(baseline_data$exclude == 1, na.rm = TRUE)
baseline_included_final <- sum(baseline_data$exclude == 0, na.rm = TRUE)

# Calculate exclusions for follow-up analysis  
followup_excluded_motion <- sum(followup_data$exclude == 1, na.rm = TRUE)
followup_included_final <- sum(followup_data$exclude == 0, na.rm = TRUE)

dairc_data_baseline_n <- nrow(dairc_data_baseline)
dairc_data_followup_n <- nrow(dairc_data_followup)

consort_exclusion <- add_box(txt = paste("All baseline & 2 year follow up records: \n(N =", format(dairc_data_full, big.mark = ","), ")")) %>%
  
  add_split(txt = c(
    paste("Baseline Participants\n(N =", format(baseline_raw_n, big.mark = ","), ")"),
    paste("2-yr Follow-up Participants\n(N =", format(followup_raw_n, big.mark = ","), ")")
  )) %>%
  
  add_box(txt = c(
    paste("Participants with Complete Demographics\nBaseline Analysis \n(N =", format(dairc_data_baseline_n, big.mark = ","), ")"),
    paste("Participants with Complete Demographics\n2-yr Follow-up Analysis \n(N =", format(dairc_data_followup_n, big.mark = ","), ")")
  )) %>%
  
    add_box(txt = c(
    paste("Complete Demographics\nBaseline Analysis DAIRC =1\n(N =", format(baseline_data_n, big.mark = ","), ")"),
    paste("Complete Demographics\n2-yr Follow-up Analysis DAIRC =1\n(N =", format(followup_data_n, big.mark = ","), ")")
  )) %>%
  
  add_split(txt = list(
    c(paste("Low Motion\nIncluded\n(N =", format(baseline_included_final, big.mark = ","), ")"),
      paste("High Motion\nExcluded\n(N =", format(baseline_excluded_motion, big.mark = ","), ")")),
    c(paste("Low Motion\nIncluded\n(N =", format(followup_included_final, big.mark = ","), ")"),
      paste("High Motion\nExcluded\n(N =", format(followup_excluded_motion, big.mark = ","), ")"))
  )) %>%
  
  add_box(txt = paste("Longitudinal Sample\nBoth timepoints\n(N =", format(length(common_subject_ids), big.mark = ","), "subjects)"))

# Generate and display the plot
plot(consort_exclusion)

# JPEG
jpeg(file.path(CONFIG$output_dir, "consort_exclusion.jpg"), width = 10, height = 8, units = "in", res = 300)
print(consort_exclusion)
dev.off()

# SVG
postscript(file.path(CONFIG$output_dir, "consort_exclusion.eps"), width = 10, height = 8)
print(consort_exclusion)
dev.off()
```

## results summary:
```{r}
cat("=== Final Results Summary ===\n\n")

cat("Sample Sizes:\n")
cat("Baseline:", nrow(baseline_data), "participants\n")
cat("Follow-up:", nrow(followup_data), "participants\n") 
cat("Longitudinal:", length(common_subject_ids), "subjects (", nrow(longitudinal_data), "observations)\n")
cat("\n=== Analysis Complete ===\n")
```


## session info:
```{r}
sessionInfo()
```











